A generalization of the Bradley-Terry model for draws in chess

### 2020-09-30 inst/jebo_chess_revision2.tex
more work
write up the rest monster work

\citet{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \citet{joe1990} uses likelihood
methods to show that this model does not fit data well.  Here we note
that such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.  \citet{moul2009} consider chess draws and use
sophisticated regression techniques to analyse games in terms of not
only the result, but also the ply-count, data not used in the present
analysis.

@@

@@

>>>>

\citet{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \citet{joe1990} uses likelihood
methods to show that this model does not fit data well.  Here we note
that such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.

The seminal statistical analysis of 
\citet{moul2009} consider chess draws and use
sophisticated regression techniques to analyse games in terms of not
only the result, but also the ply-count, data not used in the present
analysis.

@@

\definecolor{DarkGreen}{RGB}{14,99,4}

\textcolor{DarkGreen}{It is reasonable to suppose that collusion of
  the type considered here is ``effective'' in the sense that
  noncollusive games are detrimental to future play in a way that
  collusively drawn games are not.  At the end of the tournament,
  Petrosian, Keres and Geller had each played a total of 18
  noncollusive matches while Fisher, Benko and Filip had played a
  total of 26 or 27 (Tal, retiring mid tournament due to illness,
  played only 21).  For each game, we count the noncollusive games
  played up to that point in the match for White and Black.  If, in
  that game, the number of noncollusive games differ by more than 1,
  we posit a ``rest monster'' whose strength helps the player who has
  played fewer noncollusive games.  Observing that the rest monster is
  active in 49 of the 105 games of the tournament, we may consider
  these 49 games and, discarding draws, test whether rested players
  actually win more often than expected; observations are given in
  Table~\ref{restmonsterfisher} .  We reject the null of rested
  players having the same win probability as their more tired
  opponents (one-sided Fisher's exact test, $p\simeq 10^{-4}$).}

\begin{table}
\textcolor{DarkGreen}{
  \caption{Results for won games in Cura\c{c}ao 1962  between
      rested and unrested players    \label{restmonsterfisher}}
  \centering
  \fbox{%
    \begin{tabular}{l|cc}\\
      & White rested & Black rested \\ \hline
      White wins  & $9$   & $ 3$   \\
      Black wins  & $0$   & $13$   \\
  \end{tabular}}
  }
\end{table}

\textcolor{DarkGreen}{The Bradley-Terry model together with a white
  player advantage, draw monster, and rest monster, can quantify the
  strength of the rest monster at about 26\% (compare the draw monster
  at 24\%).  The null of the rest monster having zero strength is
  rejected with a likelihood ratio of about $e^{11.9}$ or a $p$-value
  of about $10^{-6}$.}

@@

\subsection*{\textcolor{DarkGreen}{The ``rest monster'' is due to a suggestion by an insightful JEBO referee}.}
